---
title: "Alzheimer's Disease and its Biopsychosocial Determinants"
author: "Daniel Middendorf"
date: "1/3/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
##########################################################
# Create alzheimer set, validation set (final hold-out test set)
##########################################################

# Downloading required libraries for analysis if they have not been downloaded yet
if(!require(tidyverse)) install.packages("tidyverse")
if(!require(caret)) install.packages("caret")
if(!require(readr)) install.packages("readr")
if(!require(ggplot2)) install.packages("ggplot2")
if(!require(psych)) install.packages("psych")
if(!require(ggthemes)) install.packages("ggthemes")
if(!require(cowplot)) install.packages("cowplot")
if(!require(corrplot)) install.packages("corrplot")
if(!require(glmnet)) install.packages("glmnet")
if(!require(knitr)) install.packages("knitr")

# Loading libraries used for the following analysis
library(tidyverse)
library(caret)
library(readr)
library(ggplot2)
library(psych)
library(ggthemes)
library(cowplot)
library(corrplot)
library(glmnet)
library(knitr)

# Loading data 
dat <- read_csv("https://raw.githubusercontent.com/dsmiddendorf/Alzheimers-Disease-Capstone-Project/main/oasis_longitudinal.csv")
dat[,c("Hand")] <- NULL # Removing column including handedness as all participants were dextrals

# Change variable name for sex from "M/F" to "Sex"
names(dat)[6] <- "Sex" 

# Validation set will be 10% of the original dataset
set.seed(1, sample.kind="Rounding") # if using R 3.5 or earlier, use `set.seed(1)`
test_index <- createDataPartition(y = dat$Group, times = 1, p = 0.1, list = FALSE)
alzheimer <- dat[-test_index,]
validation <- dat[test_index,]
rm(test_index, dat) # Removing original dataset and test index
```

# 1. Introduction
Dementia is the general intellectual deterioration resulting form brain alterations caused by pathology or normal aging. However, against popular belief, excessive forms of dementia are not part of the normal aging process and should be evaluated neuropsychologically. Neurodegenerative diseases, including Alzheimer's Disease (AD) accounting for 60-80% of all dementia cases, are the most prevalent cause of such symptoms and usually occur in old age. Despite the prevalence of such diseases, no cure has developed yet and neuropharmacological drugs can only alliviate the symptoms but not the causes. As the etiology of such diseases remained unclear, the past thirty years have seen increasingly rapid advances in the field of disease correlates and predictors. The most frequently reported correlates and predictors include: Gender, Age, Education, Socioeconomic Status, and neurophysiological brain correlates (e.g., normalized whole brain volume (nWBV), eTIV (estimated intercranial volume), atlas scaling factor (ASF)). Up to now, far too little attention has been paid to the combined effects of those predictors/correlates. Hence, this paper attempts to find a well-fitting model combining the named risk factors to determine which patients might suffer from dementia. Prior to this project, patients underwent a Clinical Dementia Rating by which they were classified as either "demented" or "nondemented" and further underwent extensive medical and neuropsychological examinations. Most patients were investigated more than once; however, in regard to the aim of this project this was not taken into account in the following analyses.


## 1.1 Description of the Dataset
The dataset that will be used for our prediction includes 335 rows and 14 columns. Each row represents a examination point of an individual patient. The scale for CDR, our dependent variable, ranges from 0 to 3 with 0 indicating no dementia and 3 indicating severe cognitive decline. 

```{r , echo = F, message = F}
df <- data.frame(Name = c("Sex", "Age", "EDUC", "SES", "MMSE", "CDR", "eTIV", "nWBV", "ASF"),
                 Description = c("Gender", "Age", "Education", "Socioeconomic Status", "Mini Mental Status Examination", "Clinical Dementia Rating",
                                 "Total Intercranial Volume", "Normalized Whole Brain Volume", "Atlas Scaling Factor"))

kable(df)
```


### Clinical Dementia Rating (CDR)
>"Clinical Dementia Rating(CDR) ranges between 0 and 3 which
depends on the issues and leisure activities of the Alzheimer’s
patient. CDR is attributed with having the option to recognize
gentle hindrances, however its shortcomings incorporate the
measure of time it takes to control, its definitive dependence
on emotional evaluation, and relative powerlessness to catch
changes over time." (Kishore et al., 2020)

```{r , echo = F, message = F}
df <- data.frame(Name = c("0", "0.5", "1", "2", "3"),
                 Description = c("None", "Very Mild", "Mild", "Moderate", "Severe"))

kable(df)
```


### Mini Mental Status Examination (MMSE)
> "The Mini-Mental State Examination (MMSE) or Folstein test is a
30-point poll that is utilized broadly in clinical and inquire about
settings to quantify psychological impairment. It is usually utilized
in medication and unified wellbeing to screen for dementia. It is
likewise used to gauge the seriousness and movement of psychological
impedance and to follow the course of subjective changes
in a person after some time; in this way making it a successful
method to record a person’s reaction to treatment." (Kishore et al., 2020)

### Total Intercranial Volume (eTIV)

> "Estimated Total Intracranial Volume (eTIV) gives the volume of
void spaces in the brain. In the event that eTIV isn’t one-sided, its
difference ought to be irregular past that clarified by the genuine
intracranial volume." (Kishore et al., 2020)

### Normalized Whole Brain Volume (nWBV)
> "Normalize Whole Brain Volume(nWBV) gives the volume of
brain other than void spaces." (Kishore et al., 2020)

### Atlas Scaling Factor (ASF)
> "Atlas Scaling Factor(ASF) is directly proportional to the eTIV and
nWBV. Atlas Scaling Factor characterized as the volume-scaling
factor required to coordinate every person to the chart book target.
Since chart book standardization likens head size, the ASF ought to
be relative to TIV." (Kishore et al., 2020)



Following you can see the first six rows of the dataset used to predict dementia and a summary of those variables:

```{r , echo = F}
head(alzheimer)
describe(alzheimer[,2:9])
```

The Gender Distribution:
```{r , echo = F}
table(alzheimer$Sex)/nrow(alzheimer)
```

The number of participants:
```{r , echo = F}
length(unique(alzheimer$`Subject ID`))
```

## 1.2 Visualization of Descriptive Statistics

```{r Density Diagramms, echo = F, fig.align='center', comment = "", message = F, results='hide',fig.keep='all'}
# Create Density Diagrams for continuous variables

## Density Plot for Age
dage <- alzheimer %>% ggplot(aes(Age)) + geom_density() + theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

## Density Plot for Education
deduc <- alzheimer %>% ggplot(aes(EDUC)) + geom_density() + theme_bw() +
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank())

## Density Plot for MMSE (Mini Mental Status Examination)
dmmse <- alzheimer %>% ggplot(aes(MMSE)) + geom_density() + theme_bw() +
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank())

## Density Plot for eTIV
detiv <- alzheimer %>% ggplot(aes(eTIV)) + geom_density() + theme_bw() +
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank())

## Density Plot for nWBV
dnwbv <- alzheimer %>% ggplot(aes(nWBV)) + geom_density() + theme_bw() +
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank())

## Density Plot for ASF
dasf <- alzheimer %>% ggplot(aes(ASF)) + geom_density() + theme_bw() +
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank())

# Putting all Density Plots together in one Figure
density_plots <- plot_grid(dage, deduc, dmmse, detiv, dnwbv, dasf)
density_plots
```

# 2. Methods and Analysis



# 2.1 Methods

# 2.2 Analysis

# 3. Results

# 4. Conclusion

# 5. References
Kishore, P., Usha Kumari, C., Kumar, M. N. V. S. S., & Pavani, T. (2020). Detection and analysis of alzheimer’s disease using various machine learning algorithms. Materials Today: Proceedings. https://doi.org/10.1016/j.matpr.2020.07.645